Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-08 doc/invt.tex
invt.tex: cut the results section

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  u_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $u_i(t)$,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-deducted bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where the bin average, $\bav{v}(t)$, is defined as
%
\begin{equation}
  \bav{v}(t)
  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vav_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

%Although the above error weighting distribution, $\pmb\rho$,
%is usually flat,
%i.e. $\rho_i \equiv 1/n$,
%in practice,
%a non-flat distribution
%is sometimes more convenient in analyzing
%some multiple-bin updating schemes later on.
%
Note that the distribution, $\pmb\rho$, is arbitrary
and independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$,
although it is usually flat.\footnote{In
well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat.}
%
%From Eqs. \eqref{eq:x_def} and \eqref{eq:vav_def},
%we have
%
%\begin{equation}
%  \sum_{i = 1}^n \rho_i \, v_{*i}(t) = 0
%  .
%\label{eq:rhox_sum}
%\end{equation}
%

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \eta_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined a relative histogram
$\eta_i(t) \equiv h_i(t) /p_i$
in the last step.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Thus, we have
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \eta_{*i}(t)
  ,
  \label{eq:vt_diffeq}
\end{equation}
%
where we have defined the bin-average-deducted
histogram, $\eta_{*i}(t)$, in a similar way to
$v_{*i}(t)$ [cf. Eq. \eqref{eq:x_def}].

We now split the histogram into
a deterministic averaged part, $\langle \eta_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \eta_i(t) =
  \langle \eta_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an ``ensemble average'',
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_V}
for the deterministic part
%
\begin{align}
  \langle \eta_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%and used $\sum_{j=1}^n p_j = 1$
%on the second line.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \eta_{*i}(t) \rangle
  \equiv
  \langle \eta_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \eta_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
as white noise,
and assume its validity for long simulations.
%
The bin-average-deducted fluctuation, $\zeta_{*i}(t)$,
%%
%\begin{equation}
%  \zeta_{*i}(t)
%  \equiv
%  \zeta_i(t)
%  -
%  \sum_{ j = 1 }^n
%    \rho_j \, \zeta_j(t)
%  ,
%  \notag
%  %\label{eq:xi_def}
%\end{equation}
%%
would also be governed by white noise as is $\zeta_i(t)$, i.e.,
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
For non-white noise,
the equivalent value of $G_i$ can be approximately obtained as
$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  p_i p_j \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho1}
\end{equation}
%
We may write the updating matrix in $\phi_{ki}$'s
as\cite{bussi2006}
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
  \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \notag
\end{equation}

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
and for the bin average of any quantity, $f$, we have
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$,
and
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
which means that the Fourier transform of
a bin-average-deducted quantity always
has a vanishing first mode.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \eta}_k(t)
,
$$
or in the continuous-time form,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \eta}_k(t)
  .
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
Deducting the bin average yields
$$
  \dot {\tilde v}_{*k}(t)
  = \alpha(t) \, \lambda_k \, {\tilde \eta}_{*k}(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_{*k}(t) - {\tilde \zeta}_{*k} (t) \bigr]
  ,
$$
where we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err(T)
  =
  \sum_{i, j = 1}^n \sum_{k=0}^{n-1}
   \phi_{ki} \, \phi_{kj}
    \left\langle v_{*i}(T) \, v_{*j}(T) \right\rangle
  =
  \sum_{k = 0}^{n - 1}
    \left\langle
      {\tilde v}_{*k}^2(T)
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
By Eq. \eqref{eq:ftstar},
we can start the last sum %in Eq. \eqref{eq:xt2_sum}
from $k = 1$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_{*k}^2(T)
    \right\rangle
  .
\label{eq:xt2_sum}
\end{equation}

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_{*k}^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \langle {\tilde \zeta}_{*k}(0)
\, {\tilde \zeta}_{*k}(t) \rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_{*k}(0)
            \, {\tilde \zeta}_{*k}(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
We may understand Eq. \eqref{eq:Gamma_sum}
as an attempt of modeling
a general histogram fluctuation, ${\tilde \zeta}_{*k}(t)$,
as an equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 1}^{n-1}
%    \dot y_k\bigl( q(t) \bigr) \,
%    \dot y_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  y_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

%Before addressing the optimal schedule
%for a multiple-bin scheme,
%we will first compute the equilibrium error
%under a constant schedule,
%$\alpha(t) = a_0$.
%%
%%In terms of the components, we have
%%
%\begin{align}
%  \left\langle \left|
%    {\tilde v}_{*k}^2
%  \right| \right\rangle
%  &=
%  \int_0^{T'}
%    \Gamma_k \, (\lambda_k \, a_0)^2 \,
%      e^{ 2 \, \lambda_k \, a_0 \, (t - T') }
%    \, dt
%  \notag
%  \\
%  &=
%  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
%  ,
%  \qquad
%  ( T' \to \infty )
%  .
%\label{eq:xt2_eql}
%\end{align}
%%

>>>>

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  u_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \notag
  %\label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $u_i(t)$,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-shifted bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where the bin average, $\bav{v}(t)$, is defined as
%
\begin{equation}
  \bav{v}(t)
  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vav_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
Below we will use the notations, $f_{*i}$ and $\bav{f}$,
for an arbitrary variable, $f_i$.
%
Note that the distribution, $\pmb\rho$, is arbitrary
and independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$,
although it is usually flat.\footnote{In
well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat.}

The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

%Although the above error weighting distribution, $\pmb\rho$,
%is usually flat,
%i.e. $\rho_i \equiv 1/n$,
%in practice,
%a non-flat distribution
%is sometimes more convenient in analyzing
%some multiple-bin updating schemes later on.
%
%
%From Eqs. \eqref{eq:x_def} and \eqref{eq:vav_def},
%we have
%
%\begin{equation}
%  \sum_{i = 1}^n \rho_i \, v_{*i}(t) = 0
%  .
%\label{eq:rhox_sum}
%\end{equation}
%

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \eta_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined a relative histogram
$\eta_i(t) \equiv h_i(t) /p_i$
in the last step.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \eta_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

We can split the histogram into
a deterministic averaged part, $\langle \eta_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \eta_i(t) =
  \langle \eta_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle \eta_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \eta_{*i}(t) \rangle
  \equiv
  \langle \eta_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \eta_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its bin-average-shifted version, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
For non-white noise,
the equivalent value of $G_i$ can be approximately obtained as
$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  p_i p_j \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \notag
  %\label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho1}
\end{equation}
%
We may write the updating matrix in $\phi_{ki}$'s
as\cite{bussi2006}
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
  \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \notag
\end{equation}

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
and for the bin average of any quantity, $f$, we have
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$,
so that
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
%
which means that the only the only effect
of bin average deduction is to
annihilate the first mode, $\tilde f_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \eta}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \eta}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err(T)
  =
  \sum_{i, j = 1}^n \sum_{k=0}^{n-1}
   \phi_{ki} \, \phi_{kj}
    \left\langle v_{*i}(T) \, v_{*j}(T) \right\rangle
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_k^2(T)
    \right\rangle
  ,
\notag
\end{align}
%
where we have used
Eq. \eqref{eq:eig_orthonormal_cols}
and drop the first term
by using Eq. \eqref{eq:ftstar}
in the second step.

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \bigl\langle {\tilde \zeta}_{*k}(0)
\, {\tilde \zeta}_{*k}(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_{*k}(0)
            \, {\tilde \zeta}_{*k}(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
We may understand Eq. \eqref{eq:Gamma_sum}
as an attempt of modeling
a general histogram fluctuation, ${\tilde \zeta}_{*k}(t)$,
as an equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 1}^{n-1}
%    \dot y_k\bigl( q(t) \bigr) \,
%    \dot y_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  y_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$, steps.
%
If the starting bias potential is zero, $u_i = 0$,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%
